Machine Learning for Predictive Auto-Tuning

### 2011-12-13 paper/draft1.tex
Add AT section
draft1 includes many figures

Due to power consumption and heat dissipation concerns, scientific applications
have shifted from computing platforms where performance had been primarily
driven by rises in the clock frequency of a single ''heavy-weight'' processor
(with complex out-of-order control and cache structures) to a platform with
ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
now not only relevant to computational sciences but to the development of all
computer systems: from ubiquitous consumer-facing devices (e.g. phones) to
high-end computer farms for web-scale applications (e.g. social networks).

@@

@@

The model-driven optimization approach uses analytical abstractions to model
the hardware architectures, the possible code transformations and their complex
interactions \cite{XXX}. Even though highly-accurate analytical models are
generally intractable to build, this approach has been quite successful in the
past at accelerating serial code \cite{XXX} with simplified but general
abstractions. However, large speed-ups for parallel code require more accurate
high-dimensional models and since this approach is bound by the quality and
scalability of its abstraction, it has been less suited for highly-specialized
kernels. This approach has been dominant in the compiler community \cite{XXX}.
As a result, it has generally been applied at compile-time where important
run-time characteristics such as input domains are missing. These drawbacks
render the model-driven optimization approach less attractive for
high-performance library developers.

@@

% Problems were considered invalid if the filters were larger than the images.
% but in the new setting with iheight >= 256 this never happened.
The product space in our study
includes 1600 argument combinations, but we restricted our experiments to correlations that
represented between 1 and 50 gigaflops of arithmetic.
Smaller problems do not fully utilize GPU hardware and are handled equally well by many kernel settings.
Larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
With the gigaflops restriction, the argument space
included 602 configurations with between 1 and 50 gigaflops.

For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
We did not use platform features ($c$) in our experiments, we leave the investigation
of cross-platform predictive auto-tuning for future work.

@@

The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
Our implementation space includes 16000 (XXX) points, too many to search exhaustively.
\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
based on empirical autotuning on several platforms, that achieves good
performance on a variety of GPUs from older-generation cards such as the
8600GT all the way to
current high-end cards such as the GTX 580 and C2070.
We call this configuration the {\em reference} implementation.
% XXX: what is reference?  Why was it chosen?

Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
the most promising part of the configuration space. This grid iterates over all
combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
XXX: isn't 73 prime?? explain this so-called ``grid''.

@@

Figure~\ref{fig:speedup} shows the effectiveness of empirical auto-tuning is in this setting.  Averaging across the range of problem configurations in our study,
implementations discovered by empirical auto-tuning are on average about 50\% (XXX check) faster than the reference implementation.
XXX: recalculate with simpler mean speedup formula.
On the GTX 295 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
On the GTX 480 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
XXX: Consider table for these numbers.
XXX: Put in some timings of mixed \& matched all-stars to show that indeed the reference was a good general baseline, and in fact the auto-tuned models are special-purpose.

\begin{figure}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over a good reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_ntrain_295.pdf}
\includegraphics[scale=.42]{fig_ntrain_480.pdf}
\caption{The effect of the invalid configuration score ($a$) and training set
size on simulated auto-tuning.
All candidates timed during the grid and
hill-climbing search procedures were used as training points, so the training
set sizes ranged from an average of 1500 (10 problem configurations) to 30
thousand (200 problem configurations).
Training on 10 or 25 configurations was useful (higher mean than the reference),
but not as useful as training on 50 or more configurations.
The results on the GTX 295 suggest that
moderate values of $a$ between $0.25$ and $0.75$ might be
best, but $a$ had no significant effect on the GTX 480.
}
\label{fig:fig_ntrain}
\end{figure}

@@

>>>>

Due to power consumption and heat dissipation concerns, scientific applications
have shifted from computing platforms where performance had been primarily
driven by rises in the clock frequency of a single ''heavy-weight'' processor
(with complex out-of-order control and cache structures) to a platform with
ever increasing numbers of ''light-weight'' cores. Interestingly, this shift is
now not only relevant to computational sciences but to the development of all
computer systems: from ubiquitous consumer-facing devices (e.g. smart phones) to
high-end computer farms for web-scale applications (e.g. social networks).

@@

% COMMENT: NP, great setup! -JB

\section{Auto-Tuning}

@@

The model-driven optimization approach uses analytical abstractions to model
the hardware architectures, the possible code transformations and their complex
interactions \cite{XXX}. Even though highly-accurate analytical models are
generally intractable to build, this approach has been quite successful in the
past at accelerating serial code \cite{XXX} with simplified but general
abstractions. However, large speed-ups for parallel code require more accurate
high-dimensional models and since this approach is bound by the quality and
scalability of its abstraction, it has been less suited for highly-specialized
% XXX: the bit above is confusing
kernels. This approach has been dominant in the compiler community \cite{XXX}.
As a result, it has generally been applied at compile-time where important
run-time characteristics such as input domains are missing. These drawbacks
render the model-driven optimization approach less attractive for
high-performance library developers.

% XXX: * empirical optimization: intro / advantages
% XXX: * empirical optimization: disadvantages
% XXX: * hybrid optimization (to counter disadvantages) ?
% XXX: * empirical optimization: w/ ML @ compile time (another way to “fix” empirical)
% XXX: * empirical optimization: ML @ runtime, especially for libs
% XXX: * empirical optimization: ML @ runtime for GPU libs why ? (w/ rtcg jit, rel. to pycuda, etc)

% XXX: our contributions
% XXX: WIIFY
% XXX: organisation of the paper

@@

The product space in our study
includes 1600 argument combinations, but we restricted our experiments to correlations that
represented between 1 and 50 gigaflops of arithmetic.
Smaller problems do not fully utilize GPU hardware and are handled equally well by many kernel settings.
Larger take so long to evaluate that there is negligible
inefficiency in implementing them via multiple calls with smaller images and fewer filters.
With the experiments searched an argument space included 602 configurations with between 1 and 50 gigaflops.

For the implementation features $b$, we directly used the integer and binary values (block\_w, block\_h, etc.) that parametrized the kernels.
We did not use platform features ($c$) in our experiments. We leave the investigation
of cross-platform predictive auto-tuning for future work.

@@

The core of any online auto-tuning algorithm (Fig.~\ref{fig:autotune_general}) is the search procedure used to explore the space of implementations.
Our implementation space includes 16000 (XXX) points, too many to search exhaustively for every input configuration.
\cite{cox+pinto:2011gcg} recommends a particular hyper-parameter implementation
based on empirical autotuning on several platforms, that achieves good
performance on a variety of GPUs from older-generation cards such as the
8600GT all the way to
current-generation flagship cards such as the GTX 580 and C2070.
We call this configuration the {\em reference} implementation.
% XXX: what is reference?  Why was it chosen?

Additionally, \cite{cox+pinto:2011gcg} advocate a particular grid search over
what was estimated to be the most relevant part of the configuration space. This grid iterates over all
combinations of XXX, but leaves the value of XXX at. In our experiments, we call this algorithm the {\em grid} search procedure.
The grid included $72$ points in addition to the reference implementation, for a total of $73$ points.
XXX: isn't 73 prime?? explain this so-called ``grid''.

@@

Figure~\ref{fig:speedup} shows the effectiveness of empirical auto-tuning is in this setting.  Averaging across the range of problem configurations in our study,
implementations discovered by empirical auto-tuning are on average about 50\% (XXX check) faster than the reference implementation.
XXX: recalculate with simpler mean speedup formula.
On the GTX 295 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
On the GTX 480 speedups over the reference implementation ranged from XXX to XXX with mean XXX.
XXX: Consider table for these numbers.
XXX: Put in some timings of mixed \& matched all-stars to show that indeed the reference was a good general baseline, and in fact the auto-tuned models are special-purpose.
\begin{figure}
\centering
\includegraphics[scale=.42]{allstars_mixup_295.pdf}
\caption{Different arguments call for different kernels: each point corresponds to a randomly drawn argument configuration.
The reference speed (horizontal) is completely uncorrelated to the speed of an optimal kernel for an unrelated argument configuration (trials performed on GTX 295).
Out of 100 random confurations, 38 (not shown) could not even be run using a kernel that was auto-tuned for different arguments.
Good performance across a variety of inputs requires {\em input-dependent auto-tuning}.
}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{speedup_295.pdf}
\includegraphics[scale=.42]{speedup_480.pdf}
\caption{Speedup of various auto-tuning strategies over the reference
implementation.
Each strategy was evaluated the same set of 83 problem configurations, which
was disjoint from the problem configurations used to build the model for
model-based hill-climbing.
The vertical dashed lines are positioned at the geometric mean speedup (XXX?)
of each strategy. The grid and hill-climbing approaches
tested 73 and 75 configurations respectively, and took 45?? XXX
seconds on average.
The model-based approach tested 0 configuration evaluations, and took 0.05
seconds on average, even with a naive Python implementation.
}
\label{fig:speedup}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_ntrain_295.pdf} % on vader
\includegraphics[scale=.42]{fig_ntrain_480.pdf} % on vader
\caption{The effect of the invalid configuration score ($a$) and training set
size on simulated auto-tuning.
All candidates timed during the grid and
hill-climbing search procedures were used as training points, so the training
set sizes ranged from an average of 1500 (10 problem configurations) to 30
thousand (200 problem configurations).
Training on 10 or 25 configurations was useful (higher mean than the reference),
but not as useful as training on 50 or more configurations.
The results on the GTX 295 suggest that
moderate values of $a$ between $0.25$ and $0.75$ might be
best, but $a$ had no significant effect on the GTX 480.
}
\label{fig:fig_ntrain}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_ntrain_munctional0_1060.pdf}
\includegraphics[scale=.42]{fig_ntrain_munctional0_2070.pdf}
\caption{more timing on munctional0}
\label{fig:fig_ntrain}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_ntrain_munctional0_480.pdf}
\includegraphics[scale=.42]{fig_ntrain_munctional0_580.pdf}
\caption{more timing on munctional0}
\label{fig:fig_ntrain}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_vader_295.pdf}
\includegraphics[scale=.42]{fig_genX_munctional0_480.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_munctional0_580.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}

\begin{figure}
\centering
\includegraphics[scale=.42]{fig_genX_munctional0_1060.pdf}
\includegraphics[scale=.42]{fig_genX_munctional0_2070.pdf}
\caption{The speedup of hill-climbing (HC) and grid algorithms for empirical autotuning}
\label{fig:speedup}
\end{figure}
